Looking for a Signal in the Noise: Revisiting Obesity and the microbiome

### 2016-03-13 Sze_ObesityMicrobiomeMeta_PeerJ_2016.Rmd
Manuscript Code Update

Age <- as.character(demographics$meanAge)

Sex <- c("111|61", "48|15", "505|2", "14|16", "112|87", "127|129", "31|27", "45|40", "Not Available")

Ancestry <- c(87.8, 0, "Not Available", 0, 100, 83, "Not Available", "Not Available", 51.4)

BMI <- c("27.0 +/- 5.3", "31.6 +/- 5.3", "26.3 +/- 4.9", "27.4 +/- 4.5", "27.7 +/- 5.3", "24.1 +/- 3.4", "24.6 +/- 4.8", "27.8 +/- 5.8", "Not Available")

Min <- c(17.5, 22.1, 16.2, 19.5, 18.2, 19, 14.0, 18.6, "Not Available")

Max <- c(46.9, 47.9, 44.8, 37.6, 51.1, 34, 41.3, 40.2, "Not Available")

table1 <- cbind(Study, Age, Sex, Ancestry, BMI, Min, Max)

colnames(table1) <- c("Study", "Age (Mean +/- SD)", "Sex (F|M)", "European Ancestry (%)", "BMI (Mean +/- SD)", "Min BMI","Max BMI")

@@

* Only BMI group information was provided and only an age range was given.

@@

```{r Data_Baxter, echo=FALSE, message=FALSE, warning=FALSE}
###########################################################################
############ Preparing Data Tables for Analysis ###########################
###########################################################################

# Reading in the necessary Data
setwd("C:/Users/marc/Desktop/obesity2/data1")
demographics <- read.csv("demographics.v2.csv")
microbiome <- read.csv("data1.subsample.otus.csv")
phylogenetic.info <- read.csv("data1.summary.taxonomy.csv")
taxonomy <- read.csv("data1.taxonomy.csv")

# Minor modifications and Rownames adustments of data tables
rownames(demographics) <- demographics[,1]
rownames(microbiome) <- microbiome[,2]
demographics <- demographics[,-1]
microbiome <- microbiome[,-2]
test.samples <- rownames(demographics)
microbiome <- microbiome[,-c(1:2)]

# Create Table for phyla information
phylogenetic.info <- phylogenetic.info[-c(1:12),]
phyla <- which(phylogenetic.info$taxlevel == 2)
phyla.table <- phylogenetic.info[phyla, ]
phyla.table <- phyla.table[,-c(1:2,4:5)]
phyla.names <- as.character(phyla.table[,1])
phyla.table <- phyla.table[,-1]
phyla.table <- as.data.frame(t(phyla.table))
colnames(phyla.table) <- phyla.names
rownames(phyla.table) <- rownames(demographics)

#Add phyla together that are not very abundant and delete them from the table
phyla.table$other <- apply(phyla.table[, c("Acidobacteria", "Deferribacteres", "Deinococcus-Thermus", "Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "Tenericutes", "Cyanobacteria_Chloroplast", "TM7")], 1, sum)
phyla.table <- phyla.table[, -c(1, 4:7, 9:11, 13, 15)]

#Create a relative abundance table for phyla
phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
phyla.table.rel.abund <- (phyla.table/phyla.total)*100
phyla.table.rel.abund <- phyla.table.rel.abund[, -8]

#Generate Alpha Diversity Measures and alpha diversity table
H <- diversity(microbiome)
S <- specnumber(microbiome)
J <- H/log(S)
alpha.diversity.shannon <- cbind(H,S,J)
alpha.test <- as.data.frame(alpha.diversity.shannon)

#Create age quartiles

demographics$age.quartile <- with(demographics, cut(Age, breaks=quantile(Age, probs=seq(0,1, by=0.25)), include.lowest=TRUE))

#Create Obese Yes/No groups
demographics$obese[demographics$BMI.classification=="Normal" | demographics$BMI.classification=="Overweight"] <- "No"
demographics$obese[demographics$BMI.classification=="Obese" | demographics$BMI.classification=="Extreme Obesity"] <- "Yes"

#Create Obese.num groups
demographics$obese.num[demographics$obese=="No"] <- 0
demographics$obese.num[demographics$obese=="Yes"] <- 1
obese <- factor(demographics$obese.num)

######################################################################################## First Level Analysis & Alpha Diversity with BMI #############
###########################################################################

#Create a column with obese and extreme obese as single entity
demographics$BMIclass2[demographics$BMI.classification=="Normal"] <- "Normal"
demographics$BMIclass2[demographics$BMI.classification=="Overweight"] <- "Overweight"                     
demographics$BMIclass2[demographics$BMI.classification=="Obese" | demographics$BMI.classification=="Extreme Obesity"] <- "Obese"

##Test BMI versus alpha diversity and phyla

bmi <- demographics$BMI

baxterH <- anova(lm(H ~ obese)) #P-value=0.01643
baxterS <- anova(lm(S ~ obese)) #P-value=0.01571
baxterJ <- anova(lm(J ~ obese)) #P-value=0.03241

#B and F tests against obesity
bacter <- phyla.table.rel.abund$Bacteroidetes
firm <- phyla.table.rel.abund$Firmicutes
BFratio <- bacter/firm

baxterBacter <- anova(lm(bacter ~ obese)) #P-value=0.3175
baxterFirm <- anova(lm(firm ~ obese)) #P-value=0.6817
baxterBF <- anova(lm(BFratio ~ obese)) #P-value=0.6305

###########################################################################
############ NMDS and PERMANOVA Analysis###################################
###########################################################################

set.seed(3)
baxter2 <- adonis(microbiome ~ obese, permutations=1000)
baxterPERM <- baxter2$aov.tab
#Not significant PERMANOVA = 0.06394, pseudo-F = 1.4246

###########################################################################
############ Relative Risk#################################################
###########################################################################

#Generate median values and put them into existing alpha.test dataframe

#Shannon diversity
alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})

##Run the RR for Shannon Diversity
H.cat <- alpha.test$shannon.cat
bmi.cat <- as.character(obese)
test3 <- cbind(H.cat, bmi.cat)
test3 <- test3[order(H.cat), ]
orderedHCat <- test3[, 1]
BaxtotalN <- length(orderedHCat)
BaxHHighTotal <- length(orderedHCat[orderedHCat=="higher"])
BaxHighShannonGroup <- as.data.frame(table(test3[c(1:BaxHHighTotal), 2]))
BaxLowShannonGroup <- as.data.frame(table(test3[c((BaxHHighTotal + 1):BaxtotalN), 2]))
#Group1 (Higher than median), obese = 22 and non-obese = 64
#Group2 (Lower than median), obese = 25 and non-obese = 61

group1 <- c(BaxHighShannonGroup[2, 2], BaxHighShannonGroup[1, 2])
group2 <- c(BaxLowShannonGroup[2, 2], BaxLowShannonGroup[1, 2])
r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group2", "group1")

baxterHEpi <- epi.2by2(r.test, method="cohort.count")
baxterHMassoc <- baxterHEpi$massoc
baxterHRR <- baxterHMassoc$RR.strata.score
baxterHRRsig <- baxterHMassoc$chisq.strata
## Risk Ratio = 1.14
## CI = 0.70, 1.85
## p-value = 0.608

##Run the RR for B/F ratio
Bacter = phyla.table.rel.abund$Bacteroidetes
Firm = phyla.table.rel.abund$Firmicutes
BFRatio = Bacter/Firm
BFRatio <- as.data.frame(BFRatio)

BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})

BFRatio.cat <- BFRatio$BFRatio.cat
test4 <- cbind(BFRatio.cat, obese)
test4 <- test4[order(BFRatio.cat), ]
orderedBFCat <- test4[, 1]
BaxBFHighTotal <- length(orderedBFCat[orderedBFCat=="higher"])
BaxHighBFGroup <- as.data.frame(table(test4[c(1:BaxBFHighTotal), 2]))
BaxLowBFGroup <- as.data.frame(table(test4[c((BaxBFHighTotal + 1):BaxtotalN), 2]))
#Group1 (Higher than median), obese = 23 and non-obese = 63
#Group2 (Lower than median), obese = 24 and non-obese = 62

group1 <- c(BaxHighBFGroup[2, 2], BaxHighBFGroup[1, 2])
group2 <- c(BaxLowBFGroup[2, 2], BaxLowBFGroup[1, 2])
r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group2", "group1")

baxterBFEpi <- epi.2by2(r.test, method="cohort.count")
baxterBFMassoc <- baxterBFEpi$massoc
baxterBFRR <- baxterBFMassoc$RR.strata.score
baxterBFRRsig <- baxterBFMassoc$chisq.strata
## Risk Ratio = 1.04
## CI = 0.64, 1.70
## p-value = 0.864

###########################################################################
############ Classification using AUCRF ###################################
###########################################################################

#generate test set
  # get rid of values that only have 0 and something else
testset <- Filter(function(x)(length(unique(x))>2), microbiome)
  # get rid of those with 0 and only 4 other values
testset <- Filter(function(x)(length(unique(x))>5), microbiome)

#Need to add phyla and alpha diversity measures to the dataset

testset <- cbind(testset, H, S, J, phyla.table.rel.abund)
testset <- cbind(obese, testset)

#Try AUCRF with default measures provided in readme
#set.seed(3)
#baxterAUCfit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)

###########################################################################
############ Z-score Data Preparation ###################################
###########################################################################

BaxterZH <- scale(H)
BaxterZLogBF <- scale(log(BFratio))
BaxterBMI <- bmi

```{r Data_Ross, echo=FALSE, message=FALSE, warning=FALSE}
###########################################################################
############ Preparing Data Tables for Analysis ###########################
###########################################################################

setwd("C:/users/marc/Desktop/obesity2/hispanic/")

hispanic.microb <- read.table("hispanic.subsample.shared", header=T)

metadata <- read.csv("s40168-015-0072-y-s1.csv")
sample.match <- read.csv("Hispanic_dataset.csv")

#Create a microbiome data table with sample names from metadata
test <- cbind(as.character(sample.match$Run_s), as.character(sample.match$Library_Name_s))
test <- test[order(test[, 1]), ]

test2 <- cbind(test, hispanic.microb)
rownames(test2) <- test2[, 2]
his.microb.edit <- test2[, -c(1:5)]
edit.metadata <- metadata[, -c(2:5)]
rownames(edit.metadata) <- edit.metadata[, 1]
edit.metadata <- edit.metadata[, -1]

#Create a metadata file in the order of the microbiome data
order1 <- rownames(his.microb.edit)
edit.metadata2 <- edit.metadata[order1, ]

#Get alpha diversity of the samples
H <- diversity(his.microb.edit)
S <- specnumber(his.microb.edit)
J <- H/log(S)
alpha.diversity.shannon <- cbind(H,S,J)
alpha.test <- as.data.frame(alpha.diversity.shannon)

#Get phyla information
#Edited out non phyla information first with sed in linux
#combined new labels with previous taxonomy file with excel
phylogenetic.info <- read.csv("phyla.data.csv")
rownames(phylogenetic.info) <- phylogenetic.info[,1]
phylogenetic.info <- phylogenetic.info[,-c(1)]
phyla.names <- as.character(phylogenetic.info$Taxonomy)
keep <- colnames(his.microb.edit)
phyla.good <- phylogenetic.info[keep, ]
phyla.names <- as.character(phyla.good[,2])
phyla.table <- his.microb.edit
colnames(phyla.table) <- phyla.names
  #add all the same columns up and then return the sum
testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
phyla.table <- as.data.frame(testing)
rm(testing)
  #combine phyla that are not that abundant
phyla.table$other <- apply(phyla.table[, c("Deinococcus-Thermus", "Elusimicrobia", "Fusobacteria", "Lentisphaerae", "Synergistetes", "TM7")], 1, sum)
phyla.table <- phyla.table[, -c(3:4, 6:7, 9:10)]

#Create a relative abundance table for phyla
phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
phyla.table.rel.abund <- (phyla.table/phyla.total)*100
phyla.table.rel.abund <- as.data.frame(phyla.table.rel.abund[, -8])

#Create BMI groups
edit.metadata2$BMI.class[edit.metadata2$BMI<=24] <- "Normal"
edit.metadata2$BMI.class[edit.metadata2$BMI>24 & edit.metadata2$BMI<30] <- "Overweight"
edit.metadata2$BMI.class[edit.metadata2$BMI>=30 & edit.metadata2$BMI<40] <- "Obese"
edit.metadata2$BMI.class[edit.metadata2$BMI>=40] <- "Extreme Obesity"

#Create Obese Yes/No groups
edit.metadata2$obese[edit.metadata2$BMI.class=="Normal" | edit.metadata2$BMI.class=="Overweight"] <- "No"
edit.metadata2$obese[edit.metadata2$BMI.class=="Obese" | edit.metadata2$BMI.class=="Extreme Obesity"] <- "Yes"

#Create BMI class 2 groups
edit.metadata2$BMI.class2[edit.metadata2$BMI<=24] <- "Normal"
edit.metadata2$BMI.class2[edit.metadata2$BMI>24 & edit.metadata2$BMI<30] <- "Overweight"
edit.metadata2$BMI.class2[edit.metadata2$BMI>=30] <- "Obese"

#Get paitent demographics to be tested

bmi <- edit.metadata2$BMI
obese <- factor(edit.metadata2$obese)

######################################################################################## First Level Analysis & Alpha Diversity with BMI #############
###########################################################################

rossH <- anova(lm(H ~ obese)) #P-value=0.1657
rossS <- anova(lm(S ~ obese)) #P-value=0.1841
rossJ <- anova(lm(J ~ obese)) #P-value=0.2538

#B and F tests against obesity
bacter <- phyla.table.rel.abund$Bacteroidetes
firm <- phyla.table.rel.abund$Firmicutes
BFratio <- bacter/firm

rossBacter <- anova(lm(bacter ~ obese)) #P-value=0.2092
rossFirm <- anova(lm(firm ~ obese)) #P-value=0.2569
rossBF <- anova(lm(BFratio ~ obese)) #P-value=0.2508

###########################################################################
############ NMDS and PERMANOVA Analysis###################################
###########################################################################

set.seed(3)
ross2 <- adonis(his.microb.edit ~ obese, permutations=1000)
rossPERM <- ross2$aov.tab
#PERMANOVA=0.7153, pseudo-F=0.73249

###########################################################################
############ Relative Risk#################################################
###########################################################################

#Generate median values and put them into existing alpha.test dataframe
#Shannon diversity

alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})

##Shannon Diversity
H.cat <- alpha.test$shannon.cat
bmi.cat <- as.character(obese)
test3 <- cbind(H.cat, bmi.cat)
test3 <- test3[order(H.cat), ]
orderedHCat <- test3[, 1]
RosstotalN <- length(orderedHCat)
RossHHighTotal <- length(orderedHCat[orderedHCat=="higher"])
RossHighShannonGroup <- as.data.frame(table(test3[c(1:RossHHighTotal), 2]))
RossLowShannonGroup <- as.data.frame(table(test3[c((RossHHighTotal + 1):RosstotalN), 2]))
group1 <- c(RossHighShannonGroup[2, 2], RossHighShannonGroup[1, 2])
group2 <- c(RossLowShannonGroup[2, 2], RossLowShannonGroup[1, 2])
#Group1 (Higher than median), obese = 18 and non-obese = 13
#Group2 (Lower than median), obese = 20 and non-obese = 12

r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group1", "group2")

rossHEpi <- epi.2by2(r.test, method="cohort.count")
rossHMassoc <- rossHEpi$massoc
rossHRR <- rossHMassoc$RR.strata.score
rossHRRsig <- rossHMassoc$chisq.strata
## Risk Ratio = 1.08
## CI = 0.72, 1.61
## p-value = 0.719

##Run the RR for B/F ratio
Bacter = phyla.table.rel.abund$Bacteroidetes
Firm = phyla.table.rel.abund$Firmicutes
BFRatio = Bacter/Firm
BFRatio <- as.data.frame(BFRatio)

BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})

BFRatio.cat <- BFRatio$BFRatio.cat
test4 <- cbind(BFRatio.cat, obese)
test4 <- test4[order(BFRatio.cat), ]
orderedBFCat <- test4[, 1]
RossBFHighTotal <- length(orderedBFCat[orderedBFCat=="higher"])
RossHighBFGroup <- as.data.frame(table(test4[c(1:RossBFHighTotal), 2]))
RossLowBFGroup <- as.data.frame(table(test4[c((RossBFHighTotal + 1):RosstotalN), 2]))
group1 <- c(RossHighBFGroup[2, 2], RossHighBFGroup[1, 2])
group2 <- c(RossLowBFGroup[2, 2], RossLowBFGroup[1, 2])
#Group1 (Higher than median), obese = 22 and non-obese = 9
#Group2 (Lower than median), obese = 16 and non-obese = 16

r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group2", "group1")

rossBFEpi <- epi.2by2(r.test, method="cohort.count")
rossBFMassoc <- rossBFEpi$massoc
rossBFRR <- rossBFMassoc$RR.strata.score
rossBFRRsig <- rossBFMassoc$chisq.strata
## Risk Ratio = 0.70
## CI = 0.47, 1.07
## p-value = 0.089

###########################################################################
############ Classification using AUCRF ###################################
###########################################################################

#Create Obese.num group
edit.metadata2$obese.num[edit.metadata2$obese=="No"] <- 0
edit.metadata2$obese.num[edit.metadata2$obese=="Yes"] <- 1
obese <- factor(edit.metadata2$obese.num)

#generate test set
  # get rid of those with 0 and only 4 other values
testset <- Filter(function(x)(length(unique(x))>5), his.microb.edit)
testset <- cbind(obese, testset)
colnames(testset)[1] <- "obese"
testset <- cbind(testset, H, S, J, phyla.table.rel.abund)

#Try AUCRF with default measures provided in readme
#set.seed(3)
#rossAUCFit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
# list of 3 Measures, AUCopt = 0.7526

###########################################################################
############ Z-score Data Preparation ###################################
###########################################################################

RossZH <- scale(H)
RossZLogBF <- scale(log(BFratio))
RossBMI <- bmi

```{r Goodrich_data, echo=FALSE, message=FALSE, warning=FALSE}
###########################################################################
############ Preparing Data Tables for Analysis ###########################
###########################################################################

setwd("C:/users/marc/Desktop/obesity2/twinsUK")

#Read in and match metadata to microbiome data
metadata <- read.csv("TwinsUKStudy_metadata.csv")
rownames(metadata) <- metadata[, 1]
shared.data <- read.table("combined.tx.1.subsample.shared", header=T)
rownames(shared.data) <- shared.data[, 2]
microbiome <- shared.data[, -c(1:3)]
keep  <- rownames(metadata)

microbiome<- microbiome[keep, ]

rm(good.metadata, shared.data, keep)

#generate alpha diversity measures with vegan
H <- diversity(microbiome)
S <- specnumber(microbiome)
J <- H/log(S)
alpha.diversity <- as.data.frame(cbind(H, S, J))
alpha.test <- alpha.diversity

#Get phyla information
#Edited out non phyla information first with sed in linux
#combined new labels with previous taxonomy file with excel
phylogenetic.info <- read.table("phyla.txt", header=T)
rownames(phylogenetic.info) <- phylogenetic.info[,1]
phylogenetic.info <- phylogenetic.info[,-c(1)]
phyla.names <- as.character(phylogenetic.info$Taxonomy)
keep <- colnames(microbiome)
phyla.good <- phylogenetic.info[keep, ]
phyla.names <- as.character(phyla.good[,2])
phyla.table <- microbiome
colnames(phyla.table) <- phyla.names
#add all the same columns up and then return the sum
testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
phyla.table <- as.data.frame(testing)
rm(testing)
#combine phyla that are not that abundant
phyla.table$other <- apply(phyla.table[, c("Acidobacteria", "Chloroflexi", "Deferribacteres", "Elusimicrobia", "Fusobacteria", "Gemmatimonadetes", "Lentisphaerae", "OD1", "Planctomycetes", "Spirochaetes", "SR1", "Synergistetes", "Tenericutes")], 1, sum)
phyla.table <- phyla.table[, -c(1, 4:6, 8:12, 14:17)]

#Create a relative abundance table for phyla
phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
phyla.table.rel.abund <- (phyla.table/phyla.total)*100

#Create BMI groups
metadata$BMI.class[metadata$body_mass_index_s<=24] <- "Normal"
metadata$BMI.class[metadata$body_mass_index_s>24 & metadata$body_mass_index_s<30] <- "Overweight"
metadata$BMI.class[metadata$body_mass_index_s>=30 & metadata$body_mass_index_s<40] <- "Obese"
metadata$BMI.class[metadata$body_mass_index_s>=40] <- "Extreme Obesity"

#Create Obese Yes/No groups
metadata$obese[metadata$BMI.class=="Normal" | metadata$BMI.class=="Overweight"] <- "No"
metadata$obese[metadata$BMI.class=="Obese" | metadata$BMI.class=="Extreme Obesity"] <- "Yes"

#Create a column with obese and extreme obese as single entity
metadata$BMIclass2[metadata$BMI.class=="Normal"] <- "Normal"
metadata$BMIclass2[metadata$BMI.class=="Overweight"] <- "Overweight"                     
metadata$BMIclass2[metadata$BMI.class=="Obese" | metadata$BMI.class=="Extreme Obesity"] <- "Obese"

#Get paitent demographics to be tested
bmi <- metadata$body_mass_index_s
obese <- factor(metadata$obese)

###########################################################################
######### First Level Analysis & Alpha Diversity with BMI #################
###########################################################################

##Test BMI versus alpha diversity and phyla

goodrichH <- anova(lm(H ~ obese)) #P-value=0.6671
goodrichS <- anova(lm(S ~ obese)) #P-value=0.9073
goodrichJ <- anova(lm(J ~ obese)) #P-value=0.6319

#B and F tests against obesity
bacter <- phyla.table.rel.abund$Bacteroidetes
firm <- phyla.table.rel.abund$Firmicutes
BFratio <- bacter/firm

goodrichBacter <- anova(lm(bacter ~ obese)) #P-value=0.6661
goodrichFirm <- anova(lm(firm ~ obese)) #P-value=0.9688
goodrichBF <- anova(lm(BFratio ~ obese)) #P-value=0.8709

###########################################################################
############ NMDS and PERMANOVA Analysis###################################
###########################################################################

set.seed(3)
goodrich2 <- adonis(microbiome ~ obese, permutations=1000)
goodrichPERM <- goodrich2$aov.tab
#PERMANOVA=0.004995, pseudo-F=2.9308

###########################################################################
############ Relative Risk#################################################
###########################################################################

#Generate median values and put them into existing alpha.test dataframe
#Shannon diversity

alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})

##Shannon Diversity
H.cat <- alpha.test$shannon.cat
bmi.cat <- as.character(obese)
test3 <- cbind(H.cat, bmi.cat)
test3 <- test3[order(H.cat), ]
orderedHCat <- test3[, 1]
GoodtotalN <- length(orderedHCat)
GoodHHighTotal <- length(orderedHCat[orderedHCat=="higher"])
GoodHighShannonGroup <- as.data.frame(table(test3[c(1:GoodHHighTotal), 2]))
GoodLowShannonGroup <- as.data.frame(table(test3[c((GoodHHighTotal + 1):GoodtotalN), 2]))
group1 <- c(GoodHighShannonGroup[2, 2], GoodHighShannonGroup[1, 2])
group2 <- c(GoodLowShannonGroup[2, 2], GoodLowShannonGroup[1, 2])
#Group1 (Higher than median), obese = 47 and non-obese = 206
#Group2 (Lower than median), obese = 56 and non-obese = 198

r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group2", "group1")

goodrichHEpi <- epi.2by2(r.test, method="cohort.count")
goodrichHMassoc <- goodrichHEpi$massoc
goodrichHRR <- goodrichHMassoc$RR.strata.score
goodrichHRRsig <- goodrichHMassoc$chisq.strata
## Risk Ratio = 1.19
## CI = 0.84, 1.68
## p-value = 0.332

##Run the RR for B/F ratio
Bacter = phyla.table.rel.abund$Bacteroidetes
Firm = phyla.table.rel.abund$Firmicutes
BFRatio = Bacter/Firm
BFRatio <- as.data.frame(BFRatio)

BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})

BFRatio.cat <- BFRatio$BFRatio.cat
test4 <- cbind(BFRatio.cat, obese)
test4 <- test4[order(BFRatio.cat), ]
orderedBFCat <- test4[, 1]
GoodBFHighTotal <- length(orderedBFCat[orderedBFCat=="higher"])
GoodHighBFGroup <- as.data.frame(table(test4[c(1:GoodBFHighTotal), 2]))
GoodLowBFGroup <- as.data.frame(table(test4[c((GoodBFHighTotal + 1):GoodtotalN), 2]))
group1 <- c(GoodHighBFGroup[2, 2], GoodHighBFGroup[1, 2])
group2 <- c(GoodLowBFGroup[2, 2], GoodLowBFGroup[1, 2])
#Group1 (Higher than median), obese = 52 and non-obese = 201
#Group2 (Lower than median), obese = 51 and non-obese = 203

r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group2", "group1")

goodrichBFEpi <- epi.2by2(r.test, method="cohort.count")
goodrichBFMassoc <- goodrichBFEpi$massoc
goodrichBFRR <- goodrichBFMassoc$RR.strata.score
goodrichBFRRsig <- goodrichBFMassoc$chisq.strata
## Risk Ratio = 0.98
## CI = 0.69, 1.38
## p-value = 0.894

###########################################################################
############ Classification using AUCRF####################################
###########################################################################

#Create Obese.num group
metadata$obese.num[metadata$obese=="No"] <- 0
metadata$obese.num[metadata$obese=="Yes"] <- 1
obese <- factor(metadata$obese.num)

#generate test set
# get rid of those with 0 and only 4 other values
testset <- Filter(function(x)(length(unique(x))>5), microbiome)
testset <- cbind(obese, testset)
colnames(testset)[1] <- "obese"
testset <- cbind(testset, H, S, J, phyla.table.rel.abund)

#Try AUCRF with default measures provided in readme
#set.seed(3)
#goodrichAUCFit <- AUCRF(obese ~ ., data=testset, ntree=1000, nodesize=20)
# list of 8 Measures, AUCopt = 0.6767159

###########################################################################
############ Z-score Data Preparation ###################################
###########################################################################

GoodrichZH <- scale(H)
GoodrichZLogBF <- scale(log(BFratio))
GoodrichBMI <- bmi

```{r escobar_data, echo=FALSE, message=FALSE, warning=FALSE}
###########################################################################
############ Preparing Data Tables for Analysis ###########################
###########################################################################

setwd("C:/users/marc/Desktop/obesity2/columbian/")

columbian.microb <- read.table("EscobarPhylotypeSub.shared", header=T)
metadata <- read.csv("columbian_dataset.csv")

#Organize the microbiome shared data
rownames(columbian.microb) <- columbian.microb[, 2]
columbian.microb <- columbian.microb[, -c(1:3)]

#Organize the metadata
rownames(metadata) <- metadata[, 5]

#Get only data that we are interested in
edit.metadata <- metadata[, -c(1:7, 9, 13:15, 16:52)]

#Sort metadata into the same order as the microbiome data
order1 <- rownames(columbian.microb)
edit.metadata2 <- edit.metadata[order1, ]

#Get alpha diversity of the samples
H <- diversity(columbian.microb)
S <- specnumber(columbian.microb)
J <- H/log(S)
alpha.diversity.shannon <- cbind(H,S,J)
alpha.test <- as.data.frame(alpha.diversity.shannon)

#Get phyla information
#Edited out non phyla information first with sed in linux
#combined new labels with previous taxonomy file with excel
phylogenetic.info <- read.table("taxonomyKey.txt")
rownames(phylogenetic.info) <- phylogenetic.info[,1]
phylogenetic.info <- phylogenetic.info[,-c(1)]
phyla.names <- as.character(phylogenetic.info$Taxonomy)
keep <- colnames(columbian.microb)
phyla.good <- phylogenetic.info[keep, ]
phyla.names <- as.character(phyla.good[,2])
phyla.table <- columbian.microb
colnames(phyla.table) <- phyla.names
#add all the same columns up and then return the sum
testing <- t(rowsum(t(phyla.table), group = rownames(t(phyla.table))))
phyla.table <- as.data.frame(testing)
rm(testing)
#combine phyla that are not that abundant
phyla.table$other <- apply(phyla.table[, c("Fusobacteria", "Lentisphaerae", "Spirochaetes", "Synergistetes", "TM7")], 1, sum)
phyla.table <- phyla.table[, -c(4:5, 7:9)]

#Create a relative abundance table for phyla
phyla.total <- apply(phyla.table[, c(1:7)], 1, sum)
phyla.table.rel.abund <- (phyla.table/phyla.total)*100
phyla.table.rel.abund <- phyla.table.rel.abund[, -8]

#Create Obese Yes/No groups
edit.metadata2$obese[edit.metadata2$description_s=="Adequate weight male stool sample" | edit.metadata2$description_s=="Overweight male stool sample" | edit.metadata2$description_s=="Adequate weight female stool sample" | edit.metadata2$description_s=="Overweight female stool sample"] <- "No"
edit.metadata2$obese[edit.metadata2$description_s=="Obese male stool sample" | edit.metadata2$description_s=="Obese female stool sample"] <- "Yes"

#Create BMI groups
edit.metadata2$BMI.class[edit.metadata2$body_mass_index_s<=24] <- "Normal"
edit.metadata2$BMI.class[edit.metadata2$body_mass_index_s>24 & edit.metadata2$body_mass_index_s<30] <- "Overweight"
edit.metadata2$BMI.class[edit.metadata2$body_mass_index_s>=30 & edit.metadata2$body_mass_index_s<40] <- "Obese"
edit.metadata2$BMI.class[edit.metadata2$body_mass_index_s>=40] <- "Extreme Obesity"

#Create BMI groups2
edit.metadata2$BMI.class2[edit.metadata2$body_mass_index_s<=24] <- "Normal"
edit.metadata2$BMI.class2[edit.metadata2$body_mass_index_s>24 & edit.metadata2$body_mass_index_s<30] <- "Overweight"
edit.metadata2$BMI.class2[edit.metadata2$body_mass_index_s>=30] <- "Obese"

#Get paitent demographics data to be tested
bmi <- edit.metadata2$body_mass_index_s
obese <- factor(edit.metadata2$obese)

######################################################################################## First Level Analysis & Alpha Diversity with BMI #############
###########################################################################

##Test BMI versus alpha diversity and phyla

escobarH <- anova(lm(H ~ obese)) #P-value=0.8636
escobarS <- anova(lm(S ~ obese)) #P-value=0.1947
escobarJ <- anova(lm(J ~ obese)) #P-value=0.4762

#B and F tests against obesity
bacter <- phyla.table.rel.abund$Bacteroidetes
firm <- phyla.table.rel.abund$Firmicutes
BFratio <- bacter/firm

escobarBacter <- anova(lm(bacter ~ obese)) #P-value=0.09253
escobarFirm <- anova(lm(firm ~ obese)) #P-value=0.3602
escobarBF <- anova(lm(BFratio ~ obese)) #P-value=0.4663

###########################################################################
############ NMDS and PERMANOVA Analysis###################################
###########################################################################

set.seed(3)
escobar2 <- adonis(columbian.microb ~ obese, permutations=1000)
escobarPERM <- escobar2$aov.tab
#PERMANOVA=0.08891, pseudo-F=1.3797

###########################################################################
############ Relative Risk#################################################
###########################################################################

#Generate median values and put them into existing alpha.test dataframe
#Shannon diversity

alpha.test <- within(alpha.test, {shannon.cat = ifelse(H <= median(H), "less", "higher")})

##Shannon Diversity
H.cat <- alpha.test$shannon.cat
bmi.cat <- as.character(obese)
test3 <- cbind(H.cat, bmi.cat)
test3 <- test3[order(H.cat), ]
orderedHCat <- test3[, 1]
EscototalN <- length(orderedHCat)
EscoHHighTotal <- length(orderedHCat[orderedHCat=="higher"])
EscoHighShannonGroup <- as.data.frame(table(test3[c(1:EscoHHighTotal), 2]))
EscoLowShannonGroup <- as.data.frame(table(test3[c((EscoHHighTotal + 1):EscototalN), 2]))
group1 <- c(EscoHighShannonGroup[2, 2], EscoHighShannonGroup[1, 2])
group2 <- c(EscoLowShannonGroup[2, 2], EscoLowShannonGroup[1, 2])
#Group1 (Higher than median), obese = 4 and non-obese = 11
#Group2 (Lower than median), obese = 6 and non-obese = 9

r.test <- rbind(group2, group1)
colnames(r.test) <- c("Obese", "Not.Obese")
rownames(r.test) <- c("group1", "group2")

escobarHEpi <- epi.2by2(r.test, method="cohort.count")
escobarHMassoc <- escobarHEpi$massoc
escobarHRR <- escobarHMassoc$RR.strata.score
escobarHRRsig <- escobarHMassoc$chisq.strata
## Risk Ratio = 1.50
## CI = 0.53, 4.26
## p-value = 0.439

##Run the RR for B/F ratio
Bacter = phyla.table.rel.abund$Bacteroidetes
Firm = phyla.table.rel.abund$Firmicutes
BFRatio = Bacter/Firm
BFRatio <- as.data.frame(BFRatio)

BFRatio <- within(BFRatio, {BFRatio.cat = ifelse(BFRatio <= median(BFRatio), "less", "higher")})

>>>>

Age <- as.character(demographics$meanAge)
AgeSD <- as.character(format(round(
  demographics$SDAge, 2), nsmall = 2))
AgeTable <- paste(Age," (",AgeSD,")", sep="")

Sex <- paste(demographics$females, "|", demographics$males)

Ancestry <- as.character(format(round(
  demographics$ancestry*100, 2), nsmall = 2))

BMI <- paste(demographics$meanBMI, " (", demographics$SDBMI, ")", sep="")

Min <- as.character(demographics$minBMI)

Max <- as.character(demographics$maxBMI)

table1 <- cbind(Study, AgeTable, Sex, Ancestry, BMI, Min, Max)
table1[table1=="NA" | table1=="NA | NA" | table1=="NA (NA)" | 
         table1=="      NA"] <- "Not Available"

colnames(table1) <- c("Study", "Age (SD)", "Sex (F|M)", "European Ancestry (%)", "BMI (SD)", "Min BMI","Max BMI")

@@

* For Turnbaugh only BMI group information was provided and only an age range was given.

@@
